---
title: "Calculations"
#author: "YOUR NAME"
date: "May 23, 2021"
output: 
  html_document:
    highlight: haddock
    theme: journal
    number_sections: no
    toc: yes
    toc_depth: 2
    toc_float: yes
    code_folding: hide
---

<style>

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  min-width: 50%;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  text-align: center;
}

</style>

  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r load_tidyverse}
# Ensure that pacman is installed for package management and loading.
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse) # for data reading wrangling and visualization

```


```{r load_packages}
# for enabling dataframe manipulation
pacman::p_load(dplyr)
# for modeling, transforming, and visualizing data
pacman::p_load(tidyverse)
# for simplifying the process of creating tidy dat
pacman::p_load(tidyr)
# for working with tabular data
pacman::p_load(data.table)
pacman::p_load(devtools)
# for data visualization
pacman::p_load(ggplot2)
# for plotting correlations 
pacman::p_load(corrplot)  # Correlations (0.84)
# provides support to ggplot2 for labeling graphs
pacman::p_load(directlabels)
# for providing a general-purpose tool for dynamic report generation
pacman::p_load(knitr)
# for providing a prettier RMarkdown (1.0.1)
pacman::p_load(kableExtra)
pacman::p_load(stargazer)
options(scipen = 1000)
pacman::p_load('fastDummies')
pacman::p_load(Formula)
source_url("https://raw.githubusercontent.com/MatthieuStigler/Misconometrics/master/Gelbach_decompo/dec_covar.R")

```


```{r, echo = FALSE}
# load cleaned CPS data
save_folder <- "intermediate"

yu_df <- readRDS(paste(save_folder, "yu_treatment_control.rds", sep="/")) 
yu_df <- yu_df %>% rename(male=sex)
yu_df <- dummy_cols(yu_df, select_columns = 'occ1digit')

# add un-normalized values for edu_num and experience
pre_proc_val <- readRDS(paste(save_folder, "pre_proc_val.rds", sep="/"))
yu_df <- yu_df %>% 
  mutate(edu_num_og = edu_num*pre_proc_val$std['edu_num']+pre_proc_val$mean['edu_num']) %>%
  mutate(experience_og = experience*pre_proc_val$std['experience']+pre_proc_val$mean['experience']) %>%
  mutate(age_og = age*pre_proc_val$std['age']+pre_proc_val$mean['age'])

```

# Alexis and Anthony

```{r,echo = FALSE}
yu_df_diff <- tidyr::spread(yu_df, treatment, log_d_earn)
names(yu_df_diff)[names(yu_df_diff)=="0"] <- "log_d_earn_counterfactual"
names(yu_df_diff)[names(yu_df_diff)=="1"] <- "log_d_earn_actual"
yu_df_diff$log_d_earn_change <- (yu_df_diff$log_d_earn_actual - yu_df_diff$log_d_earn_counterfactual)
yu_df_diff <- dummy_cols(yu_df_diff, select_columns = 'occ1digit')


yu_like_alexis <- yu_df_diff %>% 
  filter(male==0, race_black==0, age_og==18, experience_og==2, edu_num_og==12) %>% 
  select(log_d_earn_change)
paste0("Average increase in log wages over counterfactual: ", round(mean(yu_like_alexis$log_d_earn_change), 4))
paste0("Average increase in wages over counterfactual: ", (round(exp(mean(yu_like_alexis$log_d_earn_change)), 4)-1)*100, "%")

yu_like_anthony <- yu_df_diff %>% 
  filter(male==1, race_black==1, age_og==21, experience_og==5, edu_num_og==14) %>% 
  select(log_d_earn_change)
paste0("Average increase in wage growth over counterfactual for Anthonys: ", round(mean(yu_like_anthony$log_d_earn_change), 4))
paste0("Average increase in wages over counterfactual: ", (round(exp(mean(yu_like_anthony$log_d_earn_change)), 4)-1)*100, "%")

paste0("I changed experience of Anthony to 5 since no individuals in our sample had Anthony's profile (inspect test df)")
test <- yu_df_diff %>% 
  filter(male==1, race_black==1, age_og==21, edu_num_og==14) %>% 
  select(log_d_earn_change, experience_og)
```


